On the power of conditional independence testing \\ under model-X

### 2022-03-02 arxiv-v4/response_to_reviews.tex

\title{Response to reviews of\\ ``On the power of conditional independence testing \\ under model-X"\\ \vskip.2cm \Large EJS2107-037}

\author{Eugene Katsevich and Aaditya Ramdas}

\maketitle

We would like to thank the referee and the associate editor for useful comments and for the opportunity to revise our paper. Based on these suggestions, we made several improvements to the revised manuscript. We summarize these next, and then provide point-by-point responses to the comments.

@@

\section{Overview of Revisions}

To sum up the referee's main concerns, (a) the manuscripts's aim are somewhat disconnected from each other and (b) the robustness and methodological portions of the manuscript are weaker than the other portions. We agree with these assessments, and have made revisions to address both of them. Most importantly, \textbf{we restructured the paper around the main goal of studying fundamental power and optimality properties of model-X methods.} This simultaneously addresses the two main concerns by (a) by making the paper's exposition cleaner and more focused and (b) placing more emphasis on our primary contributions while placing less emphasis on our secondary contributions, the latter including the robustness and methodological aspects. In particular:

\begin{itemize}
	\item We updated Sections 1.2 and 1.3 in the introduction to reflect the focus on power and optimality of MX methods.
	\item We rewrote Section 3 to emphasize that the primary goal of the MX(2) $F$-test is a stepping stone to the power analysis of the CRT, while discussing the methodological and robustness implications as secondary consequences of independent interest.
\end{itemize}
\noindent
\textbf{Additionally, we bolstered the methodological aspect of the paper} as follows:
\begin{itemize}
	\item We clarified the relationship between the MX(2) $F$-test, the dCRT, and the GCM test in Section 3.5, including Tables 1 and 2. We discussed that the MX(2) $F$-test may be preferred to the dCRT due to its speed and that the assumptions necessary for the MX(2) $F$-test and GCM test to control Type-I error do not subsume each other. We argue that all three tests have their relative strengths and weaknesses.
	\item We conjectured that the MX(2) $F$-test continues to control the Type-I error even if the machine learning step is done in sample (Section 3.3), and updated the numerical simulation to provide support for this conjecture (Section 3.4). If true, this conjecture would make the MX(2) $F$-test more directly competitive with the dCRT and GCM tests.
\end{itemize}
\noindent 
Finally, we added several smaller revisions (primarily in Section 3) to fully address all the referee's comments. We respond point by point to these comments next.

@@

\section{Responses to Referee}

\subsection{Major comments}

\begin{enumerate}
	\item \textsl{I do find the different sections and aims of the paper a bit disconnected— specifically I’m not sure if (1) is well-connected with (2) and (3). (1) is to do with the power of the CRT test. On the other hand, (2) is about a different test that works with different assumptions, though there is an asymptotically equivalent CRT test, and (3) studies the power of this test. The motivation behind (2) seems to be concerns about the validity of the CRT assumptions (which require full knowledge of the conditional distribution $X|Z$), but I don’t see how this idea of ‘robust- ness’ really relates closely to (1). One option could be to remove or de-emphasize (1)—in my opinion it is nice to have such a result written down somewhere, but it is unsurprising.}\\

	We thank the referee for pointing out the somewhat disconnected nature of the submitted manuscript, especially the transition from Section 2 to Section 3. We agree, and we have substantially edited the exposition in Section 3 for a smoother flow. The high-level logic in the revised manuscript is that Section 2 studies the power and optimality of the CRT in finite samples against point alternatives while Sections 3 and 4 study the power of the CRT in an asymptotic setting against semiparametric alternatives. Section 3 prepares the groundwork by establishing the asymptotic equivalence between the distilled CRT and the simpler-to-analyze MX(2) $F$-test (loosening the assumption from MX to MX(2) along the way), while Section 4 establishes the power of the MX(2) $F$-test and therefore the distilled CRT. Ultimately, the central goal of the manuscript is to study the power and optimality of the CRT, and the MX(2) $F$-test is mostly a means to that end---though we argue it is of independent interest as well. This logic is conveyed in the opening paragraph of Section 3 (pp. 8-9):

@@

	\begin{quote}
		``In Section 2, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\bm Y| \bm X, \bm Z}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this ML procedure. Specifically, we consider  the power of a specific instance of the CRT against a set of semiparametric alternatives (Section 3.1). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section 3.2), which is of independent interest due to its closed form and weaker assumptions (Section 3.3). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section 3.4) and put this section's results into perspective (Section 3.5) before moving on to stating the desired power results in the next section (Section 4).''
	\end{quote}

	\item \textsl{Regarding the robustness of the CRT test, I feel one could argue that assuming exact knowledge of the first two conditional moments of $X | Z$ is not too much weaker than asking for knowledge of the entire distribution. Are there specific examples where completely correct models for the conditional means and variances are known, but not the conditional distribution? In my opinion, a more relevant analysis would quantify the impact of the error in estimating these conditional moments. Also, it might be interesting to consider a version that does not require estimating the second conditional moment, but potentially places a slightly stronger condition on the error in estimating $\mathbb E[Y | Z]$.}\\

	We agree with the referee that assuming exact knowledge of the first two moments of $X|Z$ is still a strong assumption, which we would not necessarily expect to be satisfied in practice. We nevertheless find it comforting that the MX methodologies considered here are not sensitive to misspecification of the higher moments of the assumed $X|Z$. We agree that a stronger kind of robustness analysis would quantify the impact of the error in estimating these conditional moments and/or shift more of the assumptions onto $\mathbb E[Y|Z]$, but these directions are beyond the scope of the current work. We convey these points in Section 3.5 of the revised manuscript:
	\begin{quote}
	``...Theorem 3 shows that asymptotic Type-I error control of MX-style methodologies can be achieved under the weaker MX(2) assumption, requiring only two moments of the conditional distribution $\mathcal L_n(\bm X | \bm Z)$ rather than the entire conditional distribution...note that the MX(2) assumption still requires \textit{exact} knowledge of the first and second conditional moments; we leave as an important future direction to examine the robustness of these tests to errors in these quantities. Note that bounds on the worst-case Type-I error of the CRT have been obtained in the context of a misspecified $\mathcal L_n(\bm X|\bm Z)$ (Berrett et al 2020).''
	\end{quote}

@@

\begin{quote}
``To compare our results with those in non-parametric doubly robust inference, we consider the closest representative of the latter: the generalized covariance measure (GCM) test of Shah and Peters. The GCM test statistic is defined as
\begin{equation}
	\widehat \rho^{\text{GCM}}_n \equiv \frac{1}{n}\sum_{i = 1}^n  (Y_i - \widehat g_n(Z_i))(X_i - \widehat \mu_n(Z_i)),
\end{equation}
where $\widehat \mu_n(\bm Z)$ and $\widehat g_n(\bm Z)$ are estimates of $\mu_n(\bm Z) = \mathbb E_{\mathcal L_n}[\bm X|\bm Z]$ and $g_n(\bm Z) \equiv \mathbb E_{\mathcal L_n}[\bm Y|\bm Z]$, respectively. This statistic is shown to converge under conditional independence to a mean-zero normal limit as long as the estimates of $\mathbb E_{\mathcal L_n}[\bm X|\bm Z]$ and $\mathbb E_{\mathcal L_n}[\bm Y|\bm Z]$ are both consistent, while the product in these estimation errors tends to zero at a rate of $o(n^{-1/2})$. By contrast, the MX(2) $F$-test places more weight on the model for $\bm X|\bm Z$ (assuming both first and second moments of this conditional distribution are known) while placing less weight on the model for $\bm Y|\bm Z$ (not assuming even consistency for $\mathbb E_{\mathcal L_n}[\bm Y|\bm Z]$). Therefore, while the MX(2) $F$-test closely resembles the GCM test, the assumptions required for validity of these two methods do not subsume each other (Table~\ref{tab:summary-2}).''
\end{quote}

	\begin{table}[h!]
	\centering
	\begin{tabular}{l|ll}
		Method & Guarantee & Resampling \\
		\hline
		CRT & Finite-sample & Yes \\	
		MX(2) $F$-test & Asymptotic & No  \\
		GCM test & Asymptotic & No
	\end{tabular}
	\caption{Type-I error guarantee and necessity of resampling for each method compared.}
	\label{tab:summary-1}
\end{table}

\begin{table}[h!]
	\small
	\begin{tabular}{l|lllll}
		Method & $\mathcal E(\mathbb E[\bm X|\bm Z]$) & $\mathcal E(\text{Var}[\bm X|\bm Z])$ & $\mathcal E(\mathcal L(\bm X|\bm Z))$ & $\mathcal E(\mathbb E[\bm Y|\bm Z]$) & $\mathcal E(\mathbb E[\bm X|\bm Z]) \times \mathcal E(\mathbb E[\bm Y|\bm Z])$ \\
		\hline
		CRT & 0 & 0 & 0 & --& --  \\
		MX(2) $F$-test  & 0 & 0 & -- & -- & -- \\
		GCM test & $o(1)$ & -- & -- & $o(1)$ & $o(n^{-1/2})$
	\end{tabular}
	\caption{Assumptions necessary for each method compared (excluding moment assumptions). Here, $\mathcal E(\cdot)$ refers to the root-mean-squared estimation error of a given quantity.}
	\label{tab:summary-2}
\end{table}
\end{enumerate}

\subsection{Other comments}

\begin{enumerate}
	\item \textsl{In equation (48), the RHS doesn’t depend on $j$.} \\

	While this is not a typo, the result does look strange. Intuitively, the optimal test of variable $j$ should of course depend somehow on $j$. However, as we point out in the revised manuscript, the \textit{test} (i.e. the one-bit $p$-value) does in fact depend on $j$, even if the \textit{test statistic} does not (see excerpt from p. 25 of the revision below). This is an artifact of the definition of the one-bit $p$-value of knockoffs.
	\begin{quote}
		``The reader observes that the optimal test statistic is not a function of the knockoff variables or of the index $j$, which may at first seem paradoxical. Recall from the definition (53), however, that the one-bit $p$-value compares the test statistic on the original augmented design $[X, \widetilde X]$ and its swapped version $[X, \widetilde X]_{\textnormal{swap}(j)}$. Therefore, the optimal one-bit $p$-value checks whether the original $j$th variable $X_{\bullet, j}$ fits with the rest of the data better than does its knockoff $\widetilde X_{\bullet, j}$. Therefore, the optimal one-bit $p$-value is in fact a function of the knockoffs as well as the index $j$.''
	\end{quote}

	\item \textsl{I was surprised that the non-bold $Z$ is a matrix, while the bold $Z$ is a
		vector. I would have expected it to be the other way round.}

	There are a variety of conventions in use for denoting random variables versus their realizations as well as scalars versus vectors versus matrices, and we acknowledge that our notation is inconsistent with some such conventions. However, we found that the most important notational distinction to make in our paper was between population-level random quantities and their sample-level realizations; we denoted these by bold and non-bold symbols, respectively. Our notation does not distinguish between scalars, vectors, and matrices, as the presence of subscripts disambiguates the object type. For example, $Z \in \mathbb R^{n \times p}$ is the entire matrix of covariates whereas $Z_i \in \mathbb R^{p \times 1}$ is one row of this matrix. We have added the blue text below to the notation section on p. 5 of the revision to clarify this:
	\begin{quote}
	``Recalling equations (1) and (2), population-level variables (such as $\bm X, \bm Y, \bm Z$) are denoted in boldface, while samples of these variables (such as $X_i,Y_i,Z_i$) are denoted in regular font. \textcolor{blue}{Note that boldface does \textit{not} distinguish between scalars, vectors, and matrices, as it is sometimes employed.}'' 
	\end{quote}

	\item \textsl{Where is the $X_{\bullet, j}$ etc. notation defined that is used in Section 5?} \\

	We have added an explanation of this notation near the beginning of Section 5.1 on p. 24:
	\begin{quote}
	Here, $\bm X_j \in \mathbb R$ denotes the $j$th element of the vector $\bm X \in \mathbb R^m$ and $\bm X_{-j} \in \mathbb R^{m-1}$ denotes all elements except the $j$th. Also, $X_{i,\bullet} \in \mathbb R^{n}$, $X_{\bullet, j} \in \mathbb R^m$, and $X_{\bullet, -j} \in \mathbb R^{n \times (m-1)}$ denote the $i$th row, $j$th column, and all columns but the $j$th of the matrix $X \in \mathbb R^{n \times m}$. 
	\end{quote}

### 2022-03-02 arxiv-v4/crtpower.tex
Finished pass on EJS revision

The CI testing problem under MX has benefited from several methodological innovations, but there are still many open theoretical questions, including the following three:
\begin{enumerate}
	\item[Q1.] Are there ``optimal" test statistics for MX methods, in any sense?
	\item[Q2.] To what extent can the MX assumption be weakened?
	\item[Q3.] What is the precise connection between the performance of the machine learning (ML) algorithm and the power of the resulting MX method?
\end{enumerate}
In this paper, we shed light on these questions. We summarize our findings next.

@@

While the composite nonparametric alternative of the CI problem~\eqref{conditional-independence} suggests that we cannot expect to find a uniformly most powerful test, we may still ask what is the most powerful test against a point alternative. We show that any level $\alpha$ test must also have level $\alpha$ conditionally on $(\sry, \srz)$, which allows us to reduce the composite null to a point null. We can therefore apply the Neyman-Pearson lemma to show (Section~\ref{sec:power}) that the optimal test against a point alternative $\mathcal L$ with $\mathcal L(\pry|\prx,\prz) = \bar f_{\pry|\prx,\prz}$ is the CRT based on the likelihood test statistic:
\begin{equation}
	T^{\text{opt}}(\srx; \sry, \srz) \equiv \prod_{i = 1}^n \bar f(\sry_i|\srx_i, \srz_i).
\end{equation}
The same statistic yields the most powerful one-bit $p$-values for MX knockoffs (Section~\ref{sec:knockoffs}). Despite the simplicity of the final answer, this result has not been derived before, and appears central to the design of good (powerful) test statistics. Since the model for $Y|X,Z$ is unknown, this result provides our first theoretical indication of the usefulness of ML models to learn this distribution (Q3). A3 below gives a more quantitative answer to Q3.

\paragraph{A2: The MX assumption can be drastically weakened while retaining asymptotic Type-I error control.}
Huang and Janson \cite{Huang2019} recently showed that finite-sample type-I error control is possible under only the assumption that the model for $(\prx, \prz)$ belongs to a known parametric family. We derive a complementary nonparametric result in Section~\ref{sec:weakening}, showing that if asymptotic validity suffices, then we need only the 
\begin{equation}
\begin{split}
&\textit{MX(2) assumption:} \text{ the first two moments of $\prx|\prz$ are known, i.e.} \\
&\mathbb E_{\mathcal L}[\prx|\prz] = \mu(\prz) \text{ and } \text{Var}_{\mathcal L}[\prx|\prz] = \Sigma(\prz) \text{ for known } \mu(\cdot), \Sigma(\cdot).
\label{MX(2)-intro}
\end{split}
\end{equation}
We show that the CRT, paired with the \textit{generalized covariance measure} statistic of Shah and Peters \cite{Shah2018}, retains asymptotic Type-I error control under the MX(2) assumption. Requiring knowledge of just the first two moments of the conditional distribution $\prx|\prz$, rather than the distribution itself, promises to broaden the scope of application of MX-style methodology.

\paragraph{A3: The prediction error of the ML method impacts the asymptotic efficiency of the CRT but not its consistency.}

It has been widely observed that the better the ML method approximates $\pry|\prx,\prz$, the higher power the MX method will have. We put this empirical knowledge on a theoretical foundation by expressing the asymptotic power of the CRT in terms of the prediction error of the underlying ML method (Section~\ref{sec:asymptotic-power}). In particular, we consider semiparametric alternatives of the form
\begin{equation}
H_1: \mathcal L(\pry|\prx,\prz) = N(\prx^T\beta + g(\prz),\sigma^2).
\label{parametric-alternative-intro}
\end{equation}
We analyze the power of a CRT variant that employs a separately trained estimator $\widehat g$ in an asymptotic regime where $d = \text{dim}(\prx)$ remains fixed while $p = \text{dim}(\prz)$ grows arbitrarily with the sample size $n$. We find that this test is consistent no matter what $\widehat g$ is used, while its asymptotic power against local alternatives $\beta_n = h/\sqrt{n}$ depends on the limiting mean-squared prediction error of $\widehat g$ (denoted $\mathcal E^2$) and the limiting expected variance $\mathbb E[\text{Var}[\prx | \prz]]$ (denoted $s^2$). For example, if $d = 1$, the
\begin{equation*}
\begin{split}
\text{CRT power converges to that of normal location test under alternative } N\smash{\left(\frac{hs}{\sqrt{\sigma^2 + \mathcal E^2}},1\right)}.
\end{split}
\end{equation*}
This represents the first explicit quantification of the impact of ML prediction error on the power of an MX method.
%This result clearly shows how the prediction error of $\widehat g$ impacts the asymptotic efficiency of the CRT. Unlike parallel work in causal inference, the above power result holds conditionally on $\sry, \srz$, almost surely.

@@

Recalling equations~\eqref{conditional-independence} and \eqref{eq:xyz}, population-level variables (such as $\prx,\pry,\prz$) are denoted in boldface, while samples of these variables (such as $\srx_i,\sry_i,\srz_i$) are denoted in regular font. All vectors are treated as column vectors. 
% For example, $\srx \in \mathbb R^{n \times d}, \sry \in \mathbb R^{n \times r}, \srz \in \mathbb R^{n \times p}$. 
We often use uppercase symbols to denote both random variables and their realizations (for either population- or sample-level quantities), but use lowercase to denote the latter when it is important to make this distinction. We use $\mathcal L$ to denote the joint distribution of $(\prx,\pry,\prz)$, though we sometimes use this symbol to denote the joint distribution of $(\srx,\sry,\srz)$ as well. We use the symbol ``$\equiv$" for definitions. We denote by $c_{d,1-\alpha}$ the $1-\alpha$ quantile of the $\chi^2_d$ distribution, and by $\chi^2_d(\lambda)$ the non-central $\chi^2$ distribution with $d$ degrees of freedom and noncentrality parameter $\lambda$.

@@

\paragraph*{}
Intuitively, the results of this section suggest that the more successful $\widehat f_{\pry|\prx,\prz}$ is at approximating the true alternative $f_{\pry|\prx,\prz}$, the more powerful the corresponding CRT will be. We make this relationship precise in an asymptotic setting in Section~\ref{sec:asymptotic-power}. Before examining the asymptotic power of the CRT, we establish in the next section that asymptotic Type-I error control can be obtained under a weaker form of the MX assumption.

@@

In Section~\ref{sec:power}, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\pry|\prx,\prz}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this ML procedure. Specifically, we consider  the power of a specific instance of the CRT against a set of semiparametric alternatives (Section~\ref{sec:asymptotic-setup}). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section~\ref{sec:second-order-approx}), which is of independent interest due to its closed form and weaker assumptions (Section~\ref{sec:mx2-f-test}). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section~\ref{sec:simulations}) and put this section's results into perspective (Section~\ref{sec:comparison-to-existing-results-3}) before moving on to stating the desired power results in the next section (Section~\ref{sec:asymptotic-power}).

@@

Now, consider a set of semiparametric alternatives $\mathcal L_n(\pry|\prx,\prz)$ such that
\begin{equation}
	\pry = (\prx-\mu_n(\prz))^T \beta_n + g_n(\prz) + \peps; \quad \peps \sim N(0, \sigma^2 ),\ \sigma^2 > 0
	\label{eq:linearity}
\end{equation}
for $\peps \independent (\prx, \pry, \prz)$. Here, $\beta_n \in \mathbb R^d$ is a coefficient vector, $g_n: \mathbb R^p \rightarrow \mathbb R$ a general function, and $\sigma^2 > 0$ the residual variance. Of special interest are local alternatives where $\beta_n = h/\sqrt{n}$ for some $h \in \mathbb R^d$. By Theorem~\ref{thm:power}, the most powerful test against the alternative~\eqref{eq:linearity} is the CRT based on the likelihood statistic 
\begin{equation}
T^{\text{opt}}(\srx, \sry, \srz) = \prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T \beta_n - g_n(\srz_i)\right)^2\right).
\end{equation}
Assuming local alternatives $\beta_n = h/\sqrt n$ and taking a logarithm, we obtain
\begin{equation}
	\begin{split}
		\log T^{\text{opt}}(\srx, \sry, \srz) &= -\frac n 2 \log(2\pi) - \frac{1}{2\sigma^2}\sum_{i = 1}^n\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T h/\sqrt{n} - g_n(\srz_i)\right)^2 \\
		&\approx   \frac{h^T}{\sigma^2}\frac{1}{\sqrt{n}}\sum_{i = 1}^n(\sry_i - g_n(\srz_i))(\srx_i-\mu_n(\srz_i)) + C,
	\end{split}
\label{eq:optimal-semiparametric}
\end{equation}
where $C$ is a constant that does not depend on $\srx$ and therefore does not change upon resampling.

Of course, inference based on $T^{\text{opt}}$ is infeasible because the function $g_n$ is unknown in practice. Suppose we have an estimate $\widehat g_n$ of this function. Then, the derivation~\eqref{eq:optimal-semiparametric} motivates us to base inference on the sample covariance between $\prx$ and $\pry$ after adjusting for $\prz$:
\begin{equation}
	\widehat \rho_n(\srx, \sry, \srz) \equiv \frac1n\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)).
	\label{rho-hat}
\end{equation}
For $d = 1$, the CRT rejecting for large values of $|\widehat \rho_n|$ is an instance of the \textit{distilled CRT}  (dCRT) \cite{Liu2020}, proposed for its computational speed: If $\widehat g_n$ is fit only on $(\sry, \srz)$, then it need not be refit upon resampling $\widetilde X$. The above reasoning suggests that the dCRT is also a natural test to study in the context of power against semiparametric alternatives~\eqref{eq:linearity}. In preparation to study this test's power, we extend it to $d > 1$ and propose an asymptotically equivalent test that is easier to understand.

@@

Having prepared the groundwork in Section~\ref{sec:mx2-f-test}, we present our results on the power of the dCRT against the semiparametric alternatives~\eqref{eq:linearity}. We state these results first (Section~\ref{sec:power-results}), then apply these to lasso-based dCRT (Section~\ref{sec:power-lasso-based}), and finally compare our results to existing ones (Section~\ref{sec:comparison-to-existing-results-4}). All proofs are deferred to Appendix~\ref{sec:proofs-sec45}.

>>>>

The CI testing problem under MX has benefited from several methodological innovations, but fundamental questions regarding power and optimality have received less attention. Therefore, in this paper we address the following two primary questions:
\begin{enumerate}
	\item[Q1.] Are there ``optimal" test statistics for MX methods, in any sense?
	\item[Q2.] What is the precise connection between the performance of the machine learning (ML) algorithm and the power of the resulting MX method?
\end{enumerate}
To the best of our knowledge, Q1 has not been considered before, while Q2 has only been indirectly addressed in the context of lasso-based knockoffs \cite{Weinstein2017, Liu2019, Fan2020, Weinstein2020} and CRT \cite{Wang2020b, Celentano2020}. The present paper complements these existing works by considering arbitrary ML methods. We summarize our findings next.

@@

While the composite nonparametric alternative of the CI problem~\eqref{conditional-independence} suggests that we cannot expect to find a uniformly most powerful test, we may still ask what is the most powerful test against a point alternative. We show that any level $\alpha$ test must also have level $\alpha$ conditionally on $(\sry, \srz)$, which allows us to reduce the composite null to a point null. We can therefore apply the Neyman-Pearson lemma to show (Section~\ref{sec:power}) that the optimal test against a point alternative $\mathcal L$ with $\mathcal L(\pry|\prx,\prz) = \bar f_{\pry|\prx,\prz}$ is the CRT based on the likelihood test statistic:
\begin{equation}
	T^{\text{opt}}(\srx; \sry, \srz) \equiv \prod_{i = 1}^n \bar f(\sry_i|\srx_i, \srz_i).
\end{equation}
The same statistic yields the most powerful one-bit $p$-values for MX knockoffs (Section~\ref{sec:knockoffs}). Despite the simplicity of the final answer, this result has not been derived before, and appears central to the design of good (powerful) test statistics. Since the model for $Y|X,Z$ is unknown, this result provides our first theoretical indication of the usefulness of ML models to learn this distribution (Q2). A2 below gives a more quantitative answer to Q2.

\paragraph{A2: The prediction error of the ML method impacts the asymptotic efficiency of the CRT but not its consistency.}

It has been widely observed that the better the ML method approximates $\pry|\prx,\prz$, the higher power the MX method will have. We put this empirical knowledge on a theoretical foundation by expressing the asymptotic power of the CRT in terms of the prediction error of the underlying ML method (Section~\ref{sec:asymptotic-power}). In particular, we consider semiparametric alternatives of the form
\begin{equation}
	H_1: \mathcal L(\pry|\prx,\prz) = N(\prx^T\beta + g(\prz),\sigma^2).
	\label{parametric-alternative-intro}
\end{equation}
We analyze the power of a CRT variant that employs a separately trained estimator $\widehat g$ in an asymptotic regime where $d = \text{dim}(\prx)$ remains fixed while $p = \text{dim}(\prz)$ grows arbitrarily with the sample size $n$. We find that this test is consistent no matter what $\widehat g$ is used, while its asymptotic power against local alternatives $\beta_n = h/\sqrt{n}$ depends on the limiting mean-squared prediction error of $\widehat g$ (denoted $\mathcal E^2$) and the limiting expected variance $\mathbb E[\text{Var}[\prx | \prz]]$ (denoted $s^2$). For example, if $d = 1$, the
\begin{equation*}
	\begin{split}
		\text{CRT power converges to that of normal location test under alternative } N\smash{\left(\frac{hs}{\sqrt{\sigma^2 + \mathcal E^2}},1\right)}.
	\end{split}
\end{equation*}
This represents the first explicit quantification of the impact of ML prediction error on the power of an MX method. \\

On the way to addressing Q2, we additionally establish a third result (Section~\ref{sec:weakening}) that may be of independent interest:

\paragraph{A resampling-free second-order approximation to the CRT is equivalent to the CRT and controls Type-I error under weaker assumptions.}

It was recently pointed out that if $\mathcal L(\prx|\prz)$ is Gaussian, then the resampling distribution of the dCRT test statistic can be found in closed form without actual resampling \cite{Liu2020}. Here we show that the resampling-free dCRT based on the first two moments of $\mathcal L(\prx|\prz)$ is asymptotically equivalent to the dCRT based on $\mathcal L(\prx|\prz)$ itself. Furthermore, we show the former test has asymptotic Type-I error control under the
\begin{equation}
	\begin{split}
		&\textit{MX(2) assumption:} \text{ the first two moments of $\prx|\prz$ are known, i.e.} \\
		&\mathbb E_{\mathcal L}[\prx|\prz] = \mu(\prz) \text{ and } \text{Var}_{\mathcal L}[\prx|\prz] = \Sigma(\prz) \text{ for known } \mu(\cdot), \Sigma(\cdot).
		\label{MX(2)-intro}
	\end{split}
\end{equation}
This assumption is weaker than the full MX assumption, complementing existing work \cite{Huang2019, Barber2020} on weakening assumptions for MX methods. It also suggests that the resampling-free dCRT may be used in place of the usual dCRT while achieving similar power and controlling Type-I error asymptotically.

% Huang and Jansonrecently showed that finite-sample type-I error control is possible under only the assumption that the model for $(\prx, \prz)$ belongs to a known parametric family. We derive a complementary nonparametric result in Section~\ref{sec:weakening}, showing that if asymptotic validity suffices, then we need only the

% We show that the CRT, paired with the \textit{generalized covariance measure} statistic of Shah and Peters \cite{Shah2018}, retains asymptotic Type-I error control under the MX(2) assumption. Requiring knowledge of just the first two moments of the conditional distribution $\prx|\prz$, rather than the distribution itself, promises to broaden the scope of application of MX-style methodology.

%This result clearly shows how the prediction error of $\widehat g$ impacts the asymptotic efficiency of the CRT. Unlike parallel work in causal inference, the above power result holds conditionally on $\sry, \srz$, almost surely.

@@

Recalling equations~\eqref{conditional-independence} and \eqref{eq:xyz}, population-level variables (such as $\prx,\pry,\prz$) are denoted in boldface, while samples of these variables (such as $\srx_i,\sry_i,\srz_i$) are denoted in regular font. Note that boldface does \textit{not} distinguish between scalars, vectors, and matrices, as it is sometimes employed. The dimensions of the object in this paper will be clear from context. All vectors are treated as column vectors. 
% For example, $\srx \in \mathbb R^{n \times d}, \sry \in \mathbb R^{n \times r}, \srz \in \mathbb R^{n \times p}$. 
We often use uppercase symbols to denote both random variables and their realizations (for either population- or sample-level quantities), but use lowercase to denote the latter when it is important to make this distinction. We use $\mathcal L$ to denote the joint distribution of $(\prx,\pry,\prz)$, though we sometimes use this symbol to denote the joint distribution of $(\srx,\sry,\srz)$ as well. We use the symbol ``$\equiv$" for definitions. We denote by $c_{d,1-\alpha}$ the $1-\alpha$ quantile of the $\chi^2_d$ distribution, and by $\chi^2_d(\lambda)$ the non-central $\chi^2$ distribution with $d$ degrees of freedom and noncentrality parameter $\lambda$.

@@

\paragraph*{}
Intuitively, the results of this section suggest that the more successful $\widehat f_{\pry|\prx,\prz}$ is at approximating the true alternative $f_{\pry|\prx,\prz}$, the more powerful the corresponding CRT will be. We make this relationship precise in an asymptotic setting in Section~\ref{sec:asymptotic-power}. We prepare for these results in the next section by exploring an easier-to-analyze asymptotic equivalent to the CRT.

@@

In Section~\ref{sec:power}, we saw how to construct the optimal test against point alternatives specified by $\bar f_{\pry|\prx,\prz}$. In practice, of course we do not have access to this distribution, so we usually estimate it via a statistical machine learning procedure. The goal of this section and the next is to quantitatively assess the power of the CRT as a function of the prediction error of this ML procedure. Specifically, we consider  the power of a specific instance of the CRT (the \textit{distilled CRT} \cite{Liu2020})  against a set of semiparametric alternatives (Section~\ref{sec:asymptotic-setup}). We prepare to assess the power of this test by showing its asymptotic equivalence to the simpler-to-analyze \textit{MX(2) $F$-test} (Section~\ref{sec:second-order-approx}), which is of independent interest due to its closed form and weaker assumptions (Section~\ref{sec:mx2-f-test}). We examine the finite-sample Type-I error control of the MX(2) $F$-test in numerical simulations (Section~\ref{sec:simulations}) and put this section's results into perspective (Section~\ref{sec:comparison-to-existing-results-3}) before moving on to stating the desired power results in the next section (Section~\ref{sec:asymptotic-power}).

@@

Now, consider a set of semiparametric alternatives $\mathcal L_n(\pry|\prx,\prz)$ such that
\begin{equation}
	\pry = (\prx-\mu_n(\prz))^T \beta_n + g_n(\prz) + \peps; \quad \peps \sim N(0, \sigma^2 ),\ \sigma^2 > 0
	\label{eq:linearity}
\end{equation}
for $\peps \independent (\prx, \pry, \prz)$. Here, $\beta_n \in \mathbb R^d$ is a coefficient vector, $g_n: \mathbb R^p \rightarrow \mathbb R$ a general function, and $\sigma^2 > 0$ the residual variance. Of special interest are local alternatives where $\beta_n = h/\sqrt{n}$ for some $h \in \mathbb R^d$. By Theorem~\ref{prop:crt-optimality}, the most powerful test against the alternative~\eqref{eq:linearity} is the CRT based on the likelihood statistic 
\begin{equation}
T_n^{\text{opt}}(\srx, \sry, \srz) = \prod_{i = 1}^n\frac{1}{(2\pi)^{1/2}}\exp\left(-\frac{1}{2\sigma^2}\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T \beta_n - g_n(\srz_i)\right)^2\right).
\end{equation}
Assuming local alternatives $\beta_n = h/\sqrt n$ and taking a logarithm, we obtain
\begin{equation}
	\begin{split}
		\log T_n^{\text{opt}}(\srx, \sry, \srz) &= -\frac n 2 \log(2\pi) - \frac{1}{2\sigma^2}\sum_{i = 1}^n\left(\sry_i - (\srx_i-\mu_n(\srz_i))^T h/\sqrt{n} - g_n(\srz_i)\right)^2 \\
		&\approx   \frac{h^T}{\sigma^2}\frac{1}{\sqrt{n}}\sum_{i = 1}^n(\sry_i - g_n(\srz_i))(\srx_i-\mu_n(\srz_i)) + C,
	\end{split}
\label{eq:optimal-semiparametric}
\end{equation}
where $C$ is a constant that does not depend on $\srx$ and therefore does not change upon resampling.

Of course, inference based on $T_n^{\text{opt}}$ is infeasible because the function $g_n$ is unknown in practice. Suppose we have an estimate $\widehat g_n$ of this function. Then, the derivation~\eqref{eq:optimal-semiparametric} motivates us to base inference on the sample covariance between $\prx$ and $\pry$ after adjusting for $\prz$:
\begin{equation}
	\widehat \rho_n(\srx, \sry, \srz) \equiv \frac1n\sum_{i = 1}^n (\sry_{i} - \widehat g_n(\srz_{i}))(\srx_{i} - \mu_n(\srz_i)).
	\label{rho-hat}
\end{equation}
Consider first the case $d = 1$. In this case, $\widehat \rho_n$ coincides with the recently introduced \textit{generalized covariance measure} (GCM) \cite{Shah2018}; see Section~\ref{sec:comparison-to-existing-results-3} for more a comparison to the latter work. The CRT rejecting for large values of $|\widehat \rho_n|$ is an instance of the dCRT \cite{Liu2020}, proposed for its computational speed: If $\widehat g_n$ is fit only on $(\sry, \srz)$, then it need not be refit upon resampling $\widetilde X$. The above reasoning suggests that the dCRT is also a natural test to study in the context of power against semiparametric alternatives~\eqref{eq:linearity}. In preparation to study this test's power, we extend it to $d > 1$ and propose an asymptotically equivalent test that is easier to understand.

@@

Having prepared the groundwork in Section~\ref{sec:weakening}, we present our results on the power of the dCRT against the semiparametric alternatives~\eqref{eq:linearity}. We state these results first (Section~\ref{sec:power-results}), then apply these to lasso-based dCRT (Section~\ref{sec:power-lasso-based}), and finally compare our results to existing ones (Section~\ref{sec:comparison-to-existing-results-4}). All proofs are deferred to Appendix~\ref{sec:proofs-sec45}.
